---
title: "Trajectory Differential Expression (TDE) Analysis Using Lamian"
output: html_notebook
---
## Purpose

To identify genes whose expression changes along the inferred pseudotime trajectory, we applied the Lamian framework using the trajectory differential expression (TDE) test. This analysis yields a set of genes that show significant, continuous expression changes along the posterior midgut trajectory.

### Input

* A library-size–normalized and log-transformed expression matrix comprising 13,743 genes across 1,539 WT posterior midgut cells

* Pseudotime values assigned to each cell

* Cell-level metadata, including batch information

### Workflow overview

A TDE test was performed using the Lamian framework.
The design matrix consisted of a single intercept term, with all entries set to 1, corresponding to a null model without additional covariates.
Statistical significance was assessed using 100 permutations to estimate the null distribution.

### Output

* A list of genes exhibiting significant trajectory-dependent expression changes

* 2003 genes were identified as TDE genes with an overall FDR < 0.05



```{r}
suppressMessages(library(Lamian))
vignette('Lamian')

pmg_expr <- read.csv('../../data/processed/PMG_Norm_allgenes_expr.csv', check.names = F)
head(pmg_expr)
```
```{r}
rownames(pmg_expr) <- pmg_expr[, 1]
pmg_expr
```
```{r}
pmg_expr <- pmg_expr[, -1]
head(pmg_expr)
#Lamian needs a gene by cell expression matrix
pmg_expr <- t(pmg_expr)
dim(pmg_expr)
```
```{r}
metadata <- read.csv('../../data/processed/PMG_allgenes_metadata.csv', check.names = F)
metadata
rownames(metadata) <- metadata[[1]]

metadata_f <- data.frame(row.names = rownames(metadata), cell=rownames(metadata), sample=metadata$batch) 
metadata_f
```
```{r}
pseudotime <- read.csv('../../data/processed/pseudotime_vector_pmg.csv', header = F, stringsAsFactors = F)
pseudotime <- setNames(pseudotime[[2]], pseudotime[[1]])
head(pseudotime)
```


```{r}
design <- data.frame(intercept= rep(1, 4), row.names = c('WT_01', 'WT_02', 'WT_03', 'WT_04'), group="WT")
design
```


```{r}
pmg_list <- list(expression = pmg_expr, cellanno=metadata_f, pseudotime=pseudotime, design=design)
str(pmg_list)
```
```{r}
str(pmg_list$design)
```
```{r}
summary(pmg_list$pseudotime)
anyNA(pmg_list$pseudotime)
head(names(pmg_list$pseudotime))
```


```{r}
ncol(pmg_list$design)
pmg_list$design
all(unique(pmg_list$cellanno$sample) %in% rownames(pmg_list$design))
unique(pmg_list$cellanno$sample)
```
```{r}
rownames(pmg_list$design)
```





```{r}
Res <- lamian_test(expr = pmg_list$expression,
                   cellanno = pmg_list$cellanno,
                   pseudotime = pmg_list$pseudotime,
                   design = pmg_list$design,
                   testvar = 1,
                   test.type = "time",
                   permuiter = 100, ncores = 6)
```
```{r}
head(Res$statistics)
diffgene <- rownames(Res$statistics)[Res$statistics[, 1] < 0.05]
length(diffgene)
```
```{r}
write.csv(data.frame(gene=diffgene),
          "../../data/processed/TDE_wt_pmg.csv",
          row.names = F)
```





