---
title: "Lamian XDE test - WT vs. tinman"
output: html_notebook
---
## Purpose
To identify genes that are differentially expressed between wild-type (WT) and tinman mutant cells among the set of trajectory-driving genes.

## Input
* A library-size–normalized and log-transformed expression matrix comprising 2003 trajectory-driving genes (derived from TDE analysis), measured across equal numbers of WT and tinman posterior midgut cells.

* Cell-level pseudotime values.

* Cell-level metadata, including batch and dataset information.

## Workflow overview
1. Differential trajectory analysis
A trajectory-based differential expression (XDE) test was performed using the Lamian framework.

2. Design matrix specification
The design matrix consisted of:
* an intercept term (first column, all ones), and
* a sample-level covariate representing experimental condition (WT vs tinman).

3. Statistical testing
The chi-square test was used as the primary test statistic. Due to the limited number of cells, the permutation-based test was not feasible and did not yield stable results.

## Output
* A list of genes exhibiting significant trajectory-dependent expression differences between WT and tinman.
* In total, 26 genes were identified as differentially expressed along pseudotime, with an overall FDR < 0.05.


```{r}
suppressMessages(library(Lamian))
tin_expr <- read.csv('../../data/processed/geneexpr_wt_tin.csv', check.names = F)
dim(tin_expr)
```
```{r}
rownames(tin_expr) <- tin_expr[, 1]
tin_expr
```
```{r}
tin_expr <- tin_expr[, -1]
head(tin_expr)
#Lamian needs a gene by cell expression matrix
dim(tin_expr)
tin_expr_t = t(tin_expr)
```

```{r}
metadata_tin <- read.csv('../../data/processed/metadata_wt_tin.csv', check.names = F)
metadata_tin
rownames(metadata_tin) <- metadata_tin[, 1]
metadata_tin <- metadata_tin[, -1]
head(rownames(metadata_tin))
cellanno_tin <- data.frame(
  cell   = rownames(metadata_tin),
  sample = metadata_tin$batch,
  row.names = rownames(metadata_tin),
  stringsAsFactors = FALSE
)
cellanno_tin
```
```{r}
pseudotime <- read.csv('../../data/processed/pseudotime_wt_tin.csv', header = F, stringsAsFactors = F)
pseudotime <- setNames(pseudotime[[2]], pseudotime[[1]])
head(pseudotime)
```
```{r}
design_tin <- data.frame(intercept= rep(1, 6), row.names = c('WT_01', 'WT_02', 'WT_03', 'WT_04', 'tin_01', 'tin_02'), group=c(0, 0, 0, 0, 1, 1))
design_tin
```
```{r}
list_wt_tin <- list(expression = tin_expr_t, cellanno=cellanno_tin, pseudotime=pseudotime, design=design_tin)
str(list_wt_tin)
```





```{r}
# I switched to chi-square because the permutation failed. This could be because of the small number of cells per sample.
Res_tin <- lamian_test(
  expr = list_wt_tin$expr,
  cellanno = list_wt_tin$cellanno,
  pseudotime = list_wt_tin$pseudotime,
  design = list_wt_tin$design,
  test.type = 'variable',
  testvar = 2,
  test.method = 'chisq',
  ncores = 1
)
```
```{r}
stat_tin <- Res_tin$statistics
head(stat_tin)
diffgene <- rownames(Res_tin$statistics)[Res_tin$statistics[, 1] < 0.05]
length(diffgene)
```
```{r}
write.csv(data.frame(gene=diffgene),
          "../../data/processed/XDE_wt_tin.csv",
          row.names = F)
```
```{r}
Res_tin$statistics[Res_tin$statistics$fdr.chisq.overall<0.05,]
```








